Human Factors Influencing Visual Statistical Inference

### 2015-08-28 sociology_paper.Rnw
first whole read-through

Figure~\ref{fig:location_effect} shows an overview of detection rates for both of the effects studies in experiment~9. Each dot corresponds to one lineup. The size of each dot gives the number of evaluations. Lineups with  the same set of null plots are connected by lines.  
%The detection rate for each data plot location is shown in colored by null sets. 
The overall average detection rate for each location is shown using dashed lines. Size of the points represent the number of responses. For some locations we have as many as 10 responses. For location 1, we did not have any responses for null set 1 in one of the interaction lineups.

We observe some variability in the performance due to different null sets even though the same actual data plot was used for all of  these null sets. The lineup protocol is using a finite and small sample (19 in these experiments) of all possible null sets, so it is possible to see some variability in performance depending on the null set. If null sets come from the ``extremes'' of the sampling distribution they may have structure that is more extreme than the actual data plot, making the lineup more difficult to evaluate. From the plot we can see that null set 3 appeared to be more difficult in the Interaction lineup, because the detection rate was lower regardless of the position of the actual data plot. This null set effect was tested, and a significant difference was found for null set 3 of the interaction lineup. The rest of the null sets do not show any differences at  a 1\% significance level.

@@

\begin{figure}[htbp] 
   \centering
    \includegraphics[width=6.5in]{proportion_nulls_guide.pdf} 
   \caption{Location of data plot in the lineup and proportion correct for both Interaction and Genotype effect. Each colored line represents a null set and the size of the dots represents number of responses. The overall average proportions are shown by dashed line. The actual data plot locations are shaded grey on the top panels to demonstrate their relative positions on a lineup.}
   \label{fig:location_effect}
\end{figure}

@@

Model \eqref{manova} was fit to the data to test if  detection rates were significantly different for the different locations. For this we use the {\it manova()} function provided in the {\it stats} package of the software R \citep{R:2012}. The results are shown in Table~\ref{tbl:manova}. The $p$-values for both Interaction and Genotype effect are much bigger than the conventional threshold of 0.05 suggesting that location of the actual data plot in the lineup does not significantly affect performance.

@@

Human demographics have the potential to influence performance when the lineup protocol is used for statistical inference. In our study of the demographic effects on performance across a set of different experiments we found some statistically significant effects. Age group 36-40, Countries other than India and United states, people who have a graduate degree have a significantly higher detection rate.  Gender does not have any significant effect. However, the effects are minimal, on the order of a few percentage points different from the average. This result is very important for the power of visual test as it demonstrates the robustness of the test for different human factors.

@@

Individual learning trend is observed in time taken but not so much in observer performance. Some individuals improved the performances while others showed decrease in their performances over attempts. This could be interpreted as good, in the sense that it means that observers could realistically be recruited from sources like MTurk, and that substantial training is not necessary in order to obtain useful evaluations of lineups in practice. However, we had hoped that lineups may be a useful teaching tool, to improve students ability to read data plots, and the lack of a learning trend diminishes this idea.

@@

The simulation experiment reveals that there is no significant effect of location of the actual data plot in the lineup. This is important as the visual statistical inference procedure prescribes that the data plot be placed at random in the lineup. This paper suggests that any random place in a lineup is as good as other places in the lineup. Even though there are variations on the performance depending on different null sets, their impact on probability to correctly evaluate a lineup is very negligible.

@@

There are many more ways that the lineup protocol might be tested to learn what we might expect about its performance for tackling real data mining problems. Possible changes in the lineup protocol in the pipeline are allowing observers to select more than one plot, and to vary the size of the lineup from 20 to a smaller number, and have observers all see different lineups with different null sets.  This paper assessed the effect of human factors on the experiments conducted to date. As changes to the lineup protocol are suggested by other experiments the human factor effects me need to be examined again.

>>>>

Figure~\ref{fig:location_effect} shows an overview of detection rates for both of the effects studied in experiment~9. Each dot corresponds to one lineup. The size of each dot gives the number of evaluations. Lineups with  the same set of null plots are connected by lines.  
%The detection rate for each data plot location is shown in colored by null sets. 
The overall average detection rate for each location is shown using dashed lines. Size of the points represent the number of responses. For some locations we have as many as 10 responses. For location 1, we did not have any responses for null set 1 in one of the interaction lineups.

We observe some variability in the performance due to different null sets even though the same actual data plot was used for all of  these null sets. The lineup protocol is using a finite, small sample (19 in these experiments) of all possible null sets, so it is possible to see some variability in performance depending on the null set. If null sets come from the ``extremes'' of the sampling distribution they may have structure that is more extreme than the actual data plot, making the lineup more difficult to evaluate. From the plot we can see that null set 3 appeared to be more difficult in the Interaction lineup, because the detection rate was lower regardless of the position of the actual data plot. This null set effect was tested, and a significant difference was found for null set 3 of the interaction lineup. The rest of the null sets do not show any differences at  a 1\% significance level.

@@

\begin{figure}[htbp] 
   \centering
    \includegraphics[width=6.5in]{proportion_nulls_guide.pdf} 
   \caption{Location of data plot in the lineup and proportion correct for both Interaction and Genotype effect. Each colored line represents a null set and the size of the dots represents number of responses. The overall average proportions are shown by dashed lines. The actual data plot locations are shaded grey on the top panels to demonstrate their relative positions on a lineup.}
   \label{fig:location_effect}
\end{figure}

@@

Model~\eqref{manova} was fit to the data to test, if  detection rates were significantly different for the different locations. For this we use the {\it manova()} function provided in the {\it stats} package of the software R \citep{R:2012}. The results are shown in Table~\ref{tbl:manova}. The $p$-values for both the Interaction and the Genotype effect are much bigger than the conventional threshold of 0.05 suggesting that location of the actual data plot in the lineup does not significantly affect performance.

@@

Human demographics have the potential to influence performance when the lineup protocol is used for statistical inference. In our study of the demographic effects on performance across a set of different experiments we found some statistically significant effects. Age group 36-40, Countries other than India and United states, people who have a graduate degree have a significantly higher detection rate.  Gender does not have any significant effect. However, the effects are minimal, on the order of a few percentage points different from the average. These results are very important for the power of visual tests as they demonstrate the robustness of the test against different human factors.

@@

Individual learning trend is observed in time taken but not so much in observer performance. Some individuals improved on their performances while others showed a decrease on successive attempts. This could be interpreted as good, in the sense that it means that observers could realistically be recruited from sources like MTurk, and that substantial training is not necessary in order to obtain useful evaluations of lineups in practice. However, we had hoped that lineups may be a useful teaching tool, to improve students ability to read data plots, and the lack of a learning trend diminishes this idea, at least for short term learning.

@@

The simulation experiment reveals that there is no significant effect of location on the actual data plot in the lineup. This is important as the visual statistical inference procedure prescribes that the data plot be placed at random in the lineup. This paper suggests that any random place in a lineup is as good as other places in the lineup. Even though there are variations on the performance depending on different null sets, their impact on probability to correctly evaluate a lineup is very negligible.

@@

There are many more ways that the lineup protocol might be tested to learn what we might expect about its performance for tackling real data mining problems. Possible changes in the lineup protocol in the pipeline are allowing observers to select more than one plot, and to vary the size of the lineup from 20 to a smaller number, and have observers all see different lineups with different null sets.  This paper assessed the effect of human factors on the experiments conducted to date. As changes to the lineup protocol are suggested by other experiments the human factor effects may need to be examined again.
